Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-20 Manuscript/manuscript.tex
Updates

\title{Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization} %Title of paper

@@

\begin{abstract}
  % insert abstract here
  Despite of the significant biological relevance, the binding details of cations in cellular
  lipid membranes are not fully understood. For example, consistent results for
  the binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  achieved with different experimental and theoretical methods. As recently shown in the 
  NMRlipids project, a Open Collaboration running at \url{nmrlipids.blogspot.com},
  the ion binding details could be resolved with unprecendent detail by intepreting
  the experimental NMR data with classical molecular dynamics (MD) simulations.
  However, the accuracy of the existing lipid models for MD simulations was not sufficient for this.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the simple binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
used to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

@@

While, the scaling of the partial charges improved the ion binding affinity,
it reduced the bilayer area per lipid below the experimental value. 
%Simulations with Lipid14 parameters having 
After scaling the partial charges of the head group, the glycerol
backbone and carbonyls with a factor of 0.8, the simulation gave 
the approximately 60~\AA$^2$ for the area per lipid,
which is somewhat smaller than the experimental value, 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value given by the original Lipid14 parameters, $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region,
due to the lower polarity of molecules with scaled charges. This can be compensated
by reducing the effective radius of atoms with the scaled charges by changing 
the $\sigma$ parameters in the Lennard-Jones potential, 
as previously done for ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing these $\sigma$ parameters by a factor of $f_\sigma = 0.89$, the area per molecule is restored
back close to the experimental value (see Table~\ref{tab:apls}).

@@

\subsubsection{Simulations of POPC bilayers in aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box.
As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
which is currently the most accurate three site rigid water model
\todo{More justification for the choosing the OPC3 water model are needed.
It might be good to show the comparison with the scattering data in bulk water in SI.}. 
In order to test transferability of our newly developed ECC-lipids model, 
we also performed several additional simulations with 
OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005} water models
\todo{The normal TIP3P was tested as well, right?} presented in Supporting Information (SI). 
The ECC-ion models were used for Sodium, Calcium and Chloride ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulation with the Lipid14 model was also ran with the ion model by Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ion model
was taken directly from \cite{lipid14POPC0mMNaClfiles}. \todo{Which water model was used in these simulations?} 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.

>>>>

\title{Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization} %Title of paper

@@

\begin{abstract}
  % insert abstract here
  Despite of the significant biological relevance, the binding details of cations in cellular
  lipid membranes are not fully understood. For example, consistent results for
  the binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  achieved with different experimental and theoretical methods.
  The ion binding details could be resolved with classical molecular dynamics (MD) simulations,
  however, the accuracy of the available lipid models has not been sufficient.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the ternary complex binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
used to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared spectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

@@

While, the scaling of the partial charges improved the ion binding affinity,
it reduced the bilayer area per lipid below the experimental value. 
%Simulations with Lipid14 parameters having 
After scaling the partial charges of the head group, the glycerol
backbone and carbonyls with a factor of 0.8, the simulation gave 
the approximately 60~\AA$^2$ for the area per lipid,
which is somewhat smaller than the experimental value, 64.3~\AA$^2$ \cite{kucerka11}, 
as well as the value given by the original Lipid14 parameters, $65.6 \pm 0.5$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region,
due to the lower polarity of molecules with scaled charges. This can be compensated
by reducing the effective radius of atoms with the scaled charges by changing 
the $\sigma$ parameters in the Lennard-Jones potential, 
as previously done for ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
After reducing these $\sigma$ parameters by a factor of $f_\sigma = 0.89$, the area per molecule is restored
back close to the experimental value (see Table~\ref{tab:apls}).

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of POPC bilayer in pure water or in varying salt concentrations
contained 128 POPC molecules and approximately 50~water molecules per each lipid
in the periodic orthorhombic simulation box. The water was described using 
SPC/E \cite{Berendsen1987} model. This model has the optimal dielectric constant
to be used with ECC modified molecules \todo{Citation or discussion.} and has been 
was also used in the parametrization of ECC
ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.
The results with other water models are available in SI.
%As a default, water molecules were described by the OPC3 force field~\cite{Izadi16} 
%which is currently the most accurate three site rigid water model
%In order to test transferability of our newly developed ECC-lipids model, 
%we also performed several additional simulations with 
%OPC \cite{Izadi14}, SPC/E \cite{Berendsen1987}, TIP3p-FB and TIP4p-FB~\cite{Wang2014},
%and TIP4p/2005 \cite{Abascal2005} water models
%\todo{The normal TIP3P was tested as well, right?}
%presented in Supporting Information (SI). 
The ECC-ion models \todo{Specify the exact version.} were used for Sodium, Calcium and Chloride
ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}.  % cite order: [Ca_2s, Na_s, Cl_2s]
Simulation with the Lipid14 model was also ran with the ion model by Dang et al.~\cite{smith94,chang1999,dang2006}.
Simulation data for the Lipid14 model with \AA{}qvist~\cite{aqvist90} ion model
was taken directly from \cite{lipid14POPC0mMNaClfiles}. \todo{Which water model was used in these simulations?} 
MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4).  
The simulation settings used in this work are 
summarized in Table \ref{tbl:mdpar}. 
%and they are based on previously used settings in \cite{catte16} available at \cite{lipid14POPC0mMNaClfiles}.
%\todo{As far as remember, I used there Langevin dynamics instead of thermostated MD, because
%this is done in Amber by default. If this is correct, the information in the
%table do not match with this sentence. Based on semi-extensive testing I made few years
%ago this do not change anything. Anyway, this should be reported consistently.}
Simulation trajectories and parameters are available at \cite{??} \todo{To be uploaded to Zenodo}.
